---
title: "Assessing consistency of OLS amidst omitted variable bias"
description: "EDS 241: Environmental Policy Evaluation - Lab 1"
author:
  - name: Mia Forsline
    url: {}
date: 2022-01-06
output:
  distill::distill_article:
    self_contained: false
    toc: yes
    toc_float: yes
    code_folding: show
categories: 
  - R
  - Statistics
---

# Learning Goals
- Use a simulation to demonstrate the consistency of Ordinary Least Squares (OLS) under its 3 assumptions: 
  - exogeneity: X and µ are uncorrelated
  - X and Y are independent and identically distributed 
  - large outliers in X and/or Y are rare (X and Y have finite fourth moments)
- Use a simulation to demonstrate what happens when assumptions are not upheld 
  - OLS becomes an unreliable estimator of the true population parameter $\beta_1$

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE, 
                      include = TRUE,
                      quiet = TRUE)
```

# Load necessary packages
```{r}
library(ggplot2)
library(huxtable)
library(ggthemes)
```

# Demonstrate the consistency of OLS when the assumptions are upheld 
- establish a large sample size of n = 10,000 
- generate $X_1$ and µ
- define $Y$ and the population data using a bivariate population regression function
- use a linear regression model to estimate $\beta_0$ and $\beta_1$ with an increasing sample size

```{r}
set.seed(420) 

bigN <- 10000 

X1 <- runif(bigN, min = 0, max = 10)
u <- rnorm(bigN, mean = 0, sd = 4) 


Y <- 5 + 1.5*X1 + u 
population_data <- data.frame(X1, Y) 

model1 <- lm(formula = Y ~ X1, data = population_data)
huxreg(model1, error_pos="right")

betahat_output <- matrix(ncol = 2, nrow = bigN)

for (n in 1:bigN) {
  sample <- population_data[1:n,]
  betahat_output[n,] <- lm(Y ~ X1, data = sample)$coefficients
} 

n <- seq(1,bigN)
beta1hat <- betahat_output[,c(2)]
forgraph <- data.frame(n , betahat_output[,c(2)])
```

## Graph the results
- as n increases, the OLS estimator $\hat{\beta_1}$ (blue line) approaches the true population paramater $\beta_1$ (red line) = 1.5 

```{r}
ggplot(forgraph , aes(x=n, y=beta1hat)) + geom_line(size=0.5, color="blue") +
  geom_hline(yintercept=1.5, size=2, color="red") +
  labs(x="n", y = "Beta1hat") + 
  ggthemes::theme_pander(base_size = 14) 
```

# Demonstrate the lack of OLS consistency due to omitted variables bias
- generate $X_2$, which is correlated with $X_1$
- omit $X_2$ from the model to cause omitted variable bias
- calculate $\hat{\beta_1}$ 

```{r}
X2 = X1 +rnorm(bigN , mean=0 , sd=2.2) 

Y <- 5 + 1.5*X1 + 10*X2 + u
population_data <- data.frame(X1, Y)

model1 <- lm(formula = Y ~ X1, data = population_data)
huxreg(model1, error_pos="right")

betahat_output <- matrix(ncol = 2, nrow = bigN)

for (n in 1:bigN) {
  sample <- population_data[1:n,]
  betahat_output[n,] <- lm(Y ~ X1, data = sample)$coefficients
} 

n <- seq(1,bigN)
beta1hat <- betahat_output[,c(2)]
forgraph <- data.frame(n , betahat_output[,c(2)])
```

- Compute correlation between X1 and X2, and standard deviations
- Compute "probability limit" of Beta1_hat

```{r}
cor(X1,X2)
sd(X1)
sd(X2)
1.5 + 10*cor(X1,X2)*sd(X2)/sd(X1)
```

## Graph the results 
- because of the omitted variables bias, $\hat{\beta_1}$ hate never approaches the true $\beta_1$ value even when n becomes extremely large 

```{r}
ggplot(forgraph , aes(x=n, y=beta1hat)) + geom_line(size=0.5, color="blue") +
  geom_hline(yintercept=1.5, size=2, color="red") +
  labs(x="n", y = "Beta1hat") + 
  ggthemes::theme_pander(base_size = 14) 
```

