---
title: "Robust Linear Models"
description: |
  Exploring the use of `lm_robust()` to calculate heteroskedasticity-robust standard error values 
author:
  - name: Mia Forsline
    url: {}
date: 1/20/2022
output:
  distill::distill_article:
    self_contained: no
    toc: yes
    toc_float: yes
    code_folding: show
base_url: miaforsline.github.io
slug: forsline_lm
preview: 
categories: 
  - R
  - Statistics
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE)

options(scipen=10)
```

# Learning Goals

Use `estimatr` to explore how `lm_robust()` can be used as a heteroskedasticity-robust estimator 

# Set up

-   load packages
-   import housing price data
-   visualize some of the data in a table

```{r}
library(estimatr)
library(kableExtra)
library(tidyverse)
library(modelsummary)

HPRICE2 <- read.csv("HPRICE2.csv")

kbl(head(HPRICE2)) %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

# Use `lm_robust()` to run a bivariate regression with heteroskedasticity-robust standard errors

```{r}
model1 <- lm_robust(formula = price ~ nox , data = HPRICE2)
summary(model1)
```

## Predict the housing price when `nox` = 7

-   derive the standard error and 95% confidence interval

```{r}
predicted_price <- data.frame(nox=c(7)) 
output <- predict(model1, newdata=predicted_price, se.fit=TRUE, interval='confidence')

avg_price <- round(output$fit[1], digits = 2)
ci_upper <- round(output$fit[3], digits = 2)
ci_lower <- round(output$fit[2], digits = 2)
se <- round(output$se.fit[1], digits = 2)
```

If NO~x~ = 7, we predict the average housing price to be approximately \$`r avg_price` with a 95% confidence interval of `r ci_lower` to `r ci_upper` and a standard error of `r se`.

# Use `lm_robust()` to run a multiple regression with heteroskedasticity-robust standard errors

```{r}
model2 <- lm_robust(formula = price ~ nox + rooms, data = HPRICE2)
summary(model2)
```

## Predict the housing price when `nox` = 5 and `rooms` = 6
- derive the standard error and 95% confidence interval

```{r}
predicted_price=data.frame(nox=c(5), rooms=c(6))
output <- predict(model2, newdata=predicted_price, se.fit=TRUE, interval='confidence')

avg_price <- round(output$fit[1], digits = 2)
ci_upper <- round(output$fit[3], digits = 2)
ci_lower <- round(output$fit[2], digits = 2)
se <- round(output$se.fit[1], digits = 2)
```

If NO~x~ = 5 and the house has 6 rooms, we predict the average housing price to be approximately \$`r avg_price` with a 95% confidence interval of `r ci_lower` to `r ci_upper` and a standard error of `r se`.