---
title: "Spatial analysis of 2020 Houston power outages due to extreme winter storms"
description: |
  A short description of the post.
author:
  - name: Mia Forsline
    url: {}
date: 2021-10-24
output:
  distill::distill_article:
    self_contained: no
    toc: yes
    toc_float: yes
    code_folding: show
base_url: miaforsline.github.io
slug: forsline_houston
preview: logo.png
---

# Introduction

During February 2021, Texas experienced three extreme winter storms on February 10-11, 13-17, and 15-20. These storms caused power outages in Houston's residential areas. We are interested in quantifying the severity of those power outages by calculating the number of Houston homes that lost power due to the first two storms.

To do so, we are using NASA's Worldview to examine night lights image data from February 7-16. We define an area as having experienced a blackout if we calculate a difference in night lights intensity greater than 200 nW cm^-2^ sr^-1^.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE)
```

# Load necessary packages

```{r loading packages}
library(sf)
library(stringr)
library(stars)
library(rgdal)
library(dplyr)
library(tmap)
library(lintr)
library(rosm)
```

# Function to load the DNB dataset from VNP46A1 granules

```{r}
read_dnb <- function(file_name) {

  dataset_name <- "//HDFEOS/GRIDS/VNP_Grid_DNB/Data_Fields/DNB_At_Sensor_Radiance_500m"

  h_string <- gdal_metadata(file_name)[199]
  v_string <- gdal_metadata(file_name)[219]

  tile_h <- as.integer(str_split(h_string, "=", simplify = TRUE)[[2]])
  tile_v <- as.integer(str_split(v_string, "=", simplify = TRUE)[[2]])

  west <- (10 * tile_h) - 180
  north <- 90 - (10 * tile_v)
  east <- west + 10
  south <- north - 10

  delta <- 10 / 2400

  dnb <- read_stars(file_name, sub = dataset_name, quiet = TRUE)

  st_crs(dnb) <- st_crs(4326)
  st_dimensions(dnb)$x$delta <- delta
  st_dimensions(dnb)$x$offset <- west
  st_dimensions(dnb)$y$delta <- -delta
  st_dimensions(dnb)$y$offset <- north

  return(dnb)
}
```

# Invoke the function to read in all 4 datasets

```{r}
file_name <- "data/VNP46A1.A2021038.h08v05.001.2021039064328.h5"
dnb_38_v05 <- read_dnb(file_name = file_name)

file_name <- "data/VNP46A1.A2021038.h08v06.001.2021039064329.h5"
dnb_38_v06 <- read_dnb(file_name = file_name)

file_name <- "data/VNP46A1.A2021047.h08v05.001.2021048091106.h5"
dnb_47_v05 <- read_dnb(file_name = file_name)

file_name <- "data/VNP46A1.A2021047.h08v06.001.2021048091105.h5"
dnb_47_v06 <- read_dnb(file_name = file_name)
```

# Combine the data into 2 datasets

-   `dnb_38` is data from before the storm
-   `dnb_47` is data from after the

```{r}
x1 <- dnb_38_v05
x2 <- dnb_38_v06
dnb_38 <- st_mosaic(x1, x2)

x3 <- dnb_47_v05
x4 <- dnb_47_v06
dnb_47 <- st_mosaic(x3, x4)
```

# Exploratory plots for night lights intensity before and after the storm

```{r}
plot(dnb_38,
     ylim = c(29, 30.5),
     xlim = c(-94.5, -96.5),
     breaks = seq(0, 2000, length.out = 1000))

plot(dnb_47,
     ylim = c(29, 30.5),
     xlim = c(-94.5, -96.5),
     breaks = seq(0, 2000, length.out = 1000))
```

# Create a blackout mask

-   To find the difference, we subtract night lights intensity after the storm (which is less bright) from night lights intensity before the storm (which is brighter)
-   Next, we mark all observations that did not experience a blackout as NA

```{r}
difference <- (dnb_38 - dnb_47) > 200
difference[difference == FALSE] <- NA
```

# Vectorize the mask and fix invalid geometries

```{r}
diff_vector <- st_as_sf(difference)
diff_vector <- st_make_valid(diff_vector)
```

# Crop the vectorized map to our ROI

-   Create a bounding box for the Houston area
-   Convert the bounding box to a `sf`
-   Project the bounding box to WGS 84
-   Crop the vectorized map to the Houston bounding box
-   Project the cropped, vectorized map to NAD 83

```{r}
houston <- st_polygon(list(rbind(c(-96.5, 29),
                                 c(-96.5, 30.5),
                                 c(-94.5, 30.5),
                                 c(-94.5, 29),
                                 c(-96.5, 29))))
houston <- st_sfc(houston)
houston <- houston %>%
  st_set_crs(4326)

diff_cropped <- diff_vector[houston, op = st_intersects]
diff_cropped <- diff_cropped %>%
  st_transform(3083)
```

# Roads Data

-   Use SQL query to read in the highways data
-   Project `highways` to NAD 83
-   Create and dissolve 200m buffers around the highways

```{r}
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query, quiet = TRUE)
highways <- highways %>%
  st_transform(3083)

hwy_buff_200m  <- st_buffer(highways, dist =  200)
hwy_buff_200m <- st_union(hwy_buff_200m)
```

# Buildings data

-   Use SQL query to read in the buildings data
-   Project `buildings` to NAD 83
-   Note: `buildings` contains residential, apartments, house, static_caravan, and detached types

```{r}
query <- "SELECT *
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

buildings <- st_read("data/gis_osm_buildings_a_free_1.gpkg", query = query, quiet = TRUE)
buildings <- buildings %>%
  st_transform(3083)
```

# Census tract data

-   Read in both geometries and income data
-   Extract median income field and GEOID field from `acs_income`
-   Join the datasets
-   Project the joined dataset to NAD 83
-   Project Houston bounding box to NAD 83
-   Crop the joined data to the Houston bounding box

```{r}
acs_geoms <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
                     layer = "ACS_2019_5YR_TRACT_48_TEXAS",
                     quiet = TRUE)
acs_income <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
                      layer = "X19_INCOME", 
                      quiet = TRUE)

median_income <- acs_income[c("GEOID", "B19013e1")]
geo_id <- acs_income[c("GEOID")]

geoms_income <- left_join(acs_geoms, acs_income, by = c("GEOID_Data" = "GEOID"))
geoms_income <- geoms_income %>%
  st_transform(3083)

houston <- houston %>%
  st_transform(3083)

geoms_income_cropped <- geoms_income[houston, op = st_intersects]
```

# Merge the datasets

-   Remove highway 200m buffer from the vectorized blackout mask of Houston so our area of interest ignores highways

```{r}
no_hwy <- st_difference(diff_cropped, hwy_buff_200m)
```

# Question 1: How many residential buildings were without power on 2021-02-16?

-   Use spatial subsetting to find all residential buildings in the blackout areas
-   Crop all residential buildings based on `no_hwy` (which is our area of interest)
-   Extract the number of rows (aka the number of buildings that experienced a blackout )

```{r}
buildings_blackout <- buildings[no_hwy, op = st_intersects]
q1_ans <- nrow(buildings_blackout)
```

**There were `r q1_ans` residential buildings without power on 2021-02-16.**

# Visualize buildings where the blackout occurred

-   Created a map of residential buildings that experienced a blackout (`res`)
-   Added `res` on top of a base map of the Houston area

```{r}
res <- tm_shape(buildings_blackout) +
  tm_borders() +
  tm_layout(main.title = "Residential Buildings in Houston, TX Affected by February 2021 Storm",
            main.title.size = 1, 
            main.title.position = "center")

pt1 <- st_point(c(-96.5, 29))
pt2 <- st_point(c(-96.5, 30.5))
pt3 <- st_point(c(-94.5, 30.5))
pt4 <- st_point(c(-94.5, 29))
coords <- list(rbind(pt1, pt2, pt3, pt4, pt1))
polygon <- st_polygon(x = coords)
houston <- st_sfc(polygon, crs = "EPSG:4326")

houston_bbox <- st_bbox(houston)
houston_map <- osm.raster(houston_bbox)

tm_shape(houston_map) +
tm_rgb() + 
res
```

## Merge the datasets continued 

```{r}
bb_geoms_income <- st_join(buildings_blackout, geoms_income,
                           join = st_intersects,
                           left = TRUE)
```

# Conclusion

In conclusion, we identified `r q1_ans` residential buildings in Houston without power due to the severe February 2021 storms. While night lights intensity may just be one way to measure the impact of these events, this method illustrates the vast potential of remotely sensed data from sources such as the Visible Infrared Imaging Radiometer Suite (VIIRS) aboard the Suomi satellite.