---
title: "Spatial analysis of 2020 Houston power outages due to extreme winter storms"
description: |
  EDS 223: Spatial Analysis - Assignment 2
author:
  - name: Mia Forsline
    url: {}
date: 2021-10-24
output:
  distill::distill_article:
    self_contained: no
    toc: yes
    toc_float: yes
    code_folding: show
base_url: miaforsline.github.io
slug: forsline_houston
preview: houston.jpg
---

# Learning objectives
- Load vector data
- Load raster data
- Simple raster operations
- Simple geoprocessing / vector operations
- Spatial joins
- Integrating SQL into an RMD

# Introduction

During February 2021, Texas experienced three extreme winter storms on February 10-11, February 13-17, and Februrary 15-20 that caused power outages in Houston's residential areas. We are interested in quantifying the severity of those power outages by calculating the number of Houston homes that lost power due to the first two storms.

To do so, we are used NASA's [VNP46A1](https://ladsweb.modaps.eosdis.nasa.gov/missions-and-measurements/products/VNP46A1/) data product from the [Visible Infrared Imaging Radiometer Suite (VIIRS)](https://ladsweb.modaps.eosdis.nasa.gov/missions-and-measurements/products/VNP46A1/) to examine night lights image data from February 7-16. We used four VIIRS data tiles to study the Houston area. 

We also incorporated [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84) roads and building data to account for light emitted from traffic and buildings. 

We defined an area as having experienced a blackout if we calculate a difference in night lights intensity greater than 200 nW cm^-2^ sr^-1^.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE)
```

# Load necessary packages

```{r loading packages}
library(sf)
library(stringr)
library(stars)
library(rgdal)
library(dplyr)
library(tmap)
library(lintr)
library(rosm)
```

# Load and explore data 
## Define a function to load the DNB dataset from VNP46A1 granules

```{r}
read_dnb <- function(file_name) {

  dataset_name <- "//HDFEOS/GRIDS/VNP_Grid_DNB/Data_Fields/DNB_At_Sensor_Radiance_500m"

  h_string <- gdal_metadata(file_name)[199]
  v_string <- gdal_metadata(file_name)[219]

  tile_h <- as.integer(str_split(h_string, "=", simplify = TRUE)[[2]])
  tile_v <- as.integer(str_split(v_string, "=", simplify = TRUE)[[2]])

  west <- (10 * tile_h) - 180
  north <- 90 - (10 * tile_v)
  east <- west + 10
  south <- north - 10

  delta <- 10 / 2400

  dnb <- read_stars(file_name, sub = dataset_name, quiet = TRUE)

  st_crs(dnb) <- st_crs(4326)
  st_dimensions(dnb)$x$delta <- delta
  st_dimensions(dnb)$x$offset <- west
  st_dimensions(dnb)$y$delta <- -delta
  st_dimensions(dnb)$y$offset <- north

  return(dnb)
}
```

## Invoke the function to read in all 4 datasets

```{r}
file_name <- "data/VNP46A1.A2021038.h08v05.001.2021039064328.h5"
dnb_38_v05 <- read_dnb(file_name = file_name)

file_name <- "data/VNP46A1.A2021038.h08v06.001.2021039064329.h5"
dnb_38_v06 <- read_dnb(file_name = file_name)

file_name <- "data/VNP46A1.A2021047.h08v05.001.2021048091106.h5"
dnb_47_v05 <- read_dnb(file_name = file_name)

file_name <- "data/VNP46A1.A2021047.h08v06.001.2021048091105.h5"
dnb_47_v06 <- read_dnb(file_name = file_name)
```

## Combine the data into 2 datasets

-   `dnb_38` is data from before the storm
-   `dnb_47` is data from after the storm 

```{r}
x1 <- dnb_38_v05
x2 <- dnb_38_v06
dnb_38 <- st_mosaic(x1, x2)

x3 <- dnb_47_v05
x4 <- dnb_47_v06
dnb_47 <- st_mosaic(x3, x4)
```

## Create exploratory plots for night lights intensity before and after the storm

Note: This is a crude data analysis of the "DNB_At_Sensor_Radiance_500m" dataset, which contains values as seen from space. There is no correction for stray light, moonshine, cloud presence, or any other confounding variables As a result, it is possible  we have surprising outliers and regions that systemically actually appear to be brighter from space after the storm. 

```{r before, fig.cap = "Night lights intensity in Houston, TX before the storm."}
plot(dnb_38,
     ylim = c(29, 30.5),
     xlim = c(-94.5, -96.5),
     breaks = seq(0, 2000, length.out = 1000))
```

```{r after, fig.cap = "Night lights intensity in Houston, TX after the storm."}
plot(dnb_47,
     ylim = c(29, 30.5),
     xlim = c(-94.5, -96.5),
     breaks = seq(0, 2000, length.out = 1000))
```


# Create a blackout mask

-   To find the difference, we subtract night lights intensity after the storm (which is less bright) from night lights intensity before the storm (which is brighter)
-   Next, we mark all observations that did not experience a blackout as NA

```{r}
difference <- (dnb_38 - dnb_47) > 200
difference[difference == FALSE] <- NA
```

## Vectorize the mask and fix invalid geometries

```{r}
diff_vector <- st_as_sf(difference)
diff_vector <- st_make_valid(diff_vector)
```

## Crop the vectorized map to our ROI

-   Create a bounding box for the Houston area
-   Convert the bounding box to a `sf`
-   Project the bounding box to WGS 84
-   Crop the vectorized map to the Houston bounding box
-   Project the cropped, vectorized map to NAD 83

```{r}
houston <- st_polygon(list(rbind(c(-96.5, 29),
                                 c(-96.5, 30.5),
                                 c(-94.5, 30.5),
                                 c(-94.5, 29),
                                 c(-96.5, 29))))
houston <- st_sfc(houston)
houston <- houston %>%
  st_set_crs(4326)

diff_cropped <- diff_vector[houston, op = st_intersects]
diff_cropped <- diff_cropped %>%
  st_transform(3083)
```

# Read in roads Data

-   Use SQL to read in the highways data
-   Project `highways` to NAD 83
-   Create and dissolve 200m buffers around the highways

```{r}
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query, quiet = TRUE)
highways <- highways %>%
  st_transform(3083)

hwy_buff_200m  <- st_buffer(highways, dist =  200)
hwy_buff_200m <- st_union(hwy_buff_200m)
```

# Read in buildings data

-   Use SQL to read in the buildings data
-   Project `buildings` to NAD 83
-   Note: `buildings` contains residential, apartments, house, static_caravan, and detached types

```{r}
query <- "SELECT *
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

buildings <- st_read("data/gis_osm_buildings_a_free_1.gpkg", query = query, quiet = TRUE)
buildings <- buildings %>%
  st_transform(3083)
```

# Read in census tract data

-   Read in both geometries and income data
-   Extract median income field and GEOID field from `acs_income`
-   Join the datasets
-   Project the joined dataset to NAD 83
-   Project Houston bounding box to NAD 83
-   Crop the joined data to the Houston bounding box

```{r}
acs_geoms <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
                     layer = "ACS_2019_5YR_TRACT_48_TEXAS",
                     quiet = TRUE)
acs_income <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
                      layer = "X19_INCOME", 
                      quiet = TRUE)

median_income <- acs_income[c("GEOID", "B19013e1")]
geo_id <- acs_income[c("GEOID")]

geoms_income <- left_join(acs_geoms, acs_income, by = c("GEOID_Data" = "GEOID"))
geoms_income <- geoms_income %>%
  st_transform(3083)

houston <- houston %>%
  st_transform(3083)

geoms_income_cropped <- geoms_income[houston, op = st_intersects]
```

# Merge the datasets

-   Remove highway 200m buffer from the vectorized blackout mask of Houston so our area of interest ignores highways

```{r}
no_hwy <- st_difference(diff_cropped, hwy_buff_200m)
```

# Question 1: How many residential buildings were without power on 2021-02-16?

-   Use spatial subsetting to find all residential buildings in the blackout areas
-   Crop all residential buildings based on `no_hwy` (which is our area of interest)
-   Extract the number of rows (aka the number of buildings that experienced a blackout )

```{r}
buildings_blackout <- buildings[no_hwy, op = st_intersects]
q1_ans <- nrow(buildings_blackout)
```

**There were `r q1_ans` residential buildings without power on 2021-02-16.**

## Visualize buildings where the blackout occurred

-   Created a map of residential buildings that experienced a blackout (`res`)
-   Added `res` on top of a base map of the Houston area

```{r blackout, fig.cap = "Approximately 157,411 residential buildings in Houston, TX were affected by the February 2021 storm."}
res <- tm_shape(buildings_blackout) +
  tm_borders()

pt1 <- st_point(c(-96.5, 29))
pt2 <- st_point(c(-96.5, 30.5))
pt3 <- st_point(c(-94.5, 30.5))
pt4 <- st_point(c(-94.5, 29))
coords <- list(rbind(pt1, pt2, pt3, pt4, pt1))
polygon <- st_polygon(x = coords)
houston <- st_sfc(polygon, crs = "EPSG:4326")

houston_bbox <- st_bbox(houston)
houston_map <- osm.raster(houston_bbox)

tm_shape(houston_map) +
tm_rgb() + 
res

tmap_save(filename = "houston.jpg", width = 8, height = 5, units = "in", dpi = 300)

bb_geoms_income <- st_join(buildings_blackout, geoms_income,
                           join = st_intersects,
                           left = TRUE)
```

# Conclusion

In conclusion, we identified `r q1_ans` residential buildings in Houston without power due to the severe February 2021 storms. While night lights intensity may just be one way to measure the impact of these events, this method illustrates the vast potential of remotely sensed data from sources such as the Visible Infrared Imaging Radiometer Suite (VIIRS) aboard the Suomi satellite.

# GitHub
Full code for this assignment can be found [here](https://github.com/eds223-fall-2021-org/spatial-analysis-21). 